---
title: "**Student performance**"
author: "guilinXue"
date: "2020/6/13"
output:
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE,
  warning = FALSE,
  comment = ' ',
  message = FALSE,
  collapse = TRUE,
  prompt = TRUE,
  fig.width = 9,
  fig.height = 8,
  dpi = 300,
  highlight = FALSE,
  cache = FALSE
  )
```

### Introduction
Hi Friends, welcome to my kernel! I have used dplyr and ggplot2 to manipulate and visualise the student performance data. 
I have then created a random forest to create a simple prediction model. 
If you find this script useful, please upvote or leave a comment with your feedback, Thanks! 

### Loading data and libraries
key libraries used include:

ggplot2 - Data visualisation 

dplyr - Data manipulation

kableExtra, xtable - Creating tables

RColorBrewer - Using colours

```{r  warnings = FALSE, message=FALSE} 

Data <- read.csv("StudentsPerformance.csv",header=TRUE)
library(kableExtra)
library(xtable)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyr)
```

### Understanding the data
The data has eight variables and 1000 observations. The key variables include gender, race/ethnicity, parental level of education,lunch, test preparation course, math score, reading score and writing score.

```{r message= FALSE, warning=FALSE}
str(Data)
```

### Exploratory data analysis

1. ***Students By gender***: There are more 518 female students and 482 male students. 

```{r }
plot1 <- 
  ggplot() +
  geom_bar(data = Data, aes(x = Data$gender), width = 0.2, fill = "green") +
  geom_text(stat='count', data = Data, aes(x = Data$gender, label=..count..), vjust=-0.2) +
  theme_bw() +
  xlab("Gender") +
  ylab("Number of Students") +
  theme_classic() +
  ggtitle("Number of Students by Gender") +
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill") +
  ylim(0, 600)
  
plot1

```

2. ***Students By race***: There are 319 students in group C, 262 students in group D while there are only 89 students in group A.

```{r }
plot2 <- ggplot() +
  geom_bar(data = Data, aes(x = Data$race.ethnicity), width = 0.6, fill = "green") +
  geom_text(data = Data, aes(x = Data$race.ethnicity, label = ..count..), stat = "count", vjust = -0.2) +
  theme_bw() +
  xlab("Race/Ethnicity") +
  ylab("Number of Students") +
  theme(
    text = element_text(family = "Tahoma")
  ) +
  theme_classic()+
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill") +
  ggtitle("Number of Students by Race/Ethnicity")
plot2

```

3. ***Distribution of scores***
152 students scored between 65 - 70 marks in maths. 154 students scored in the same bracket in writing. 148 students scored between 70 - 75 marks in reading.  


```{r message=FALSE}
Data_gather <- Data %>%
  mutate(StudentID = row_number()) %>%
  gather(key = "subject", value = "score", math.score:writing.score)
  
  Data_gather %>%
  ggplot(aes(x = score)) +
  geom_histogram(bins = 20, colour = "green", fill = "white")  +
  facet_grid(subject ~.) +
  theme_bw() +
  theme_classic() +
  scale_x_continuous(limits = c(0,100), breaks = seq(0, 100, 10)) +
  #geom_label(aes(label = ..count..), stat = "count", y = 150, size = 2) +
  stat_bin(bins= 20, geom="text", aes(label=..count..) , 
vjust = -1) +
  ylim(0, 200)

```

### Preparation courses 

1. ***Completion***: Majority of students do not complete preparation courses. Proportion of students completing preparation course is highest in group E. 


```{r message= FALSE, warning=FALSE}
plot3 <- ggplot() +
  geom_bar(data = Data, aes(x = Data$test.preparation.course), width = 0.6, fill = "green") +
  geom_text(data = Data, aes(x = Data$test.preparation.course, label = ..count..), stat = "count", vjust = -0.2) +
 # geom_bar(data = Data, aes(x = Data$test.preparation.course, y = race.ethnicity), fill = "#bcd4e6", width = 0.6) +
  facet_grid(. ~race.ethnicity) +
  theme_bw() +
  xlab("Preparation course") +
  ylab("Number of Students") +
  theme(
    text = element_text(family = "Tahoma") ) + 
  theme_classic() +
  ggtitle("Number of Students who completed Preparation course by Race/Ethnicity") +
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill")

plot3
```

2. **Impact of course**: As we see the that the average of the scores of students who completed preparation course is higher than the average of scores of the students who did not complete the course.
 
```{r message= FALSE, warning=FALSE}
plot4 <- ggplot(data=Data_gather, aes(x=Data_gather$test.preparation.course, y=Data_gather$score, 
                             fill=Data_gather$test.preparation.course)) + 
  geom_boxplot() +
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE) +
  theme_bw() +
  theme_classic() +
  xlab("Preparation Course") +
  ylab("Average Score") +
  theme(
    text = element_text(family = "Tahoma"),
    legend.title = element_blank()) +
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill")

plot4

```

### Scores by race/ethnicity

Average scores are highest in the group E.

```{r message= FALSE, warning=FALSE}
plot5 <- ggplot(data=Data_gather, aes(x=Data_gather$race.ethnicity, y=Data_gather$score, 
                             fill=Data_gather$race.ethnicity)) + 
  geom_boxplot() + 
    theme(
    text = element_text(family = "Tahoma"),
    axis.text.x = element_blank(),
    legend.title = element_blank()
    ) +

  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE) +
  facet_grid(. ~ subject ) +
  theme_bw() +
  theme_classic() +
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill") +
  xlab("Race/ethnicity") +
  ylab("Score")

plot5


```

### Scores By gender

Female students have high average scores in reading and writing, while male students fare better in maths.


```{r message= FALSE, warning=FALSE}

plot6 <- ggplot(data=Data_gather, aes(x=Data_gather$gender, y=Data_gather$score, 
                             fill=Data_gather$gender)) + 
  geom_boxplot() +
  stat_summary(fun.y=mean, colour="darkred", geom="point", 
               shape=18, size=3,show_guide = FALSE) +
  facet_grid(. ~ subject ) +
  theme_bw() +
  theme_classic() +
  theme(legend.title = element_blank()) +
  xlab("Gender") +
  ylab("Score") +
  scale_fill_brewer(type = "qual", palette = 1, direction = 1,
  aesthetics = "fill")

plot6
```

### Impact of the type of lunch

```{r message= FALSE, warning=FALSE}
Data_gather_mean <- Data_gather %>%
  group_by(StudentID) %>%
  summarise(mean = mean(score))


table1 <- merge(Data_gather_mean, Data_gather[,c(1:6)], by = "StudentID") %>%
  distinct(StudentID, .keep_all = TRUE) %>%
  group_by(race.ethnicity, lunch) %>%
  summarise(avg.score = mean(mean, na.rm = TRUE))

kable(table1)

```

### Correlation between reading and writing score

There is a strong correlation between reading scores and the writing scores with a coefficient of 0.95.

```{r message= FALSE, warning=FALSE}
ggplot() +
  geom_point(data = Data, aes(x = Data$math.score, y = Data$reading.score)) +
  theme_bw() +
  theme_classic() +
  xlab("Math Score") +
  ylab("Reading Score")

ggplot() +
  geom_point(data = Data, aes(x = Data$writing.score, y = Data$reading.score)) +
    theme_bw() +
  theme_classic() +
  xlab("Writing Score") +
  ylab("Reading Score")


```

```{r message= FALSE, warning=FALSE}

cor.test(Data$math.score, Data$reading.score)
cor.test(Data$math.score, Data$writing.score)
cor.test(Data$writing.score, Data$reading.score)

```

### Prediction 
 
We created a simple prediction model, to see if we predict whether the student had a test preparation course or not based on all other variables. 

```{r message= FALSE, warning=FALSE}
library(randomForest)
library(caret)
Data1 <- Data

Data1$gender <- as.numeric(factor(Data1$gender, levels = unique(Data1$gender)))
Data1$race.ethnicity <- as.numeric(factor(Data1$race.ethnicity, levels = unique(Data1$race.ethnicity)))
Data1$parental.level.of.education <- as.numeric(factor(Data1$parental.level.of.education , levels = unique(Data1$parental.level.of.education)))
Data1$lunch  <- as.numeric(factor(Data1$lunch, levels = unique(Data1$lunch)))
Data1$test.preparation.course <- as.numeric(factor(Data1$test.preparation.course, levels = unique(Data1$test.preparation.course)))
Data1$test.preparation.course = as.factor(Data1$test.preparation.course)

```

Let's divide the set into training and test set with 70% of the records in the training set and the remaining in the test set. 

```{r message= FALSE, warning=FALSE}
set.seed(125)
train <- sample(nrow(Data1), 0.7*nrow(Data1), replace = FALSE)

TrainingSet <- Data1[train,]
TestSet <- Data1[-train,]

str(TrainingSet)
str(TestSet)

```

Here is a simple random forest model that has an accuracy of 71.67%.

```{r message= FALSE, warning=FALSE}
model1 <- randomForest(test.preparation.course ~ ., data = TrainingSet, importance = TRUE, ntree = 500, mtry = 6)
model1

importance(model1)
varImpPlot(model1)

PredTest <- predict(model1, TestSet, type = "class")
mean(PredTest == TestSet$test.preparation.course)


```

If you got to this point, then don't forget to leave a comment. 
Thanks!

