---
title: "Assign3"
author: "guilinXue"
date: "2020/6/13"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE,
  warning = FALSE,
  comment = ' ',
  message = FALSE,
  collapse = TRUE,
  prompt = TRUE,
  fig.width = 9,
  fig.height = 8,
  dpi = 300,
  highlight = FALSE,
  cache = FALSE
  )
```
# 1 The first question 
Looking at the bond information, we can get the following information(Because the bond information doesn't have the price to maturity information, so I'm assuming the maturity price is 120):

Price|times|Coupon rate|par value|maturity price 
|:-:|:-:|:-:|:-:|:-:|
100|20|2.68%|100|120|


## 1.1 Yield to maturity

```{r}
Y_ma <- function(r, p, Cs, Cp){
  n <- length(Cs)
  tt <- 1:n
  Yi <- p - sum(Cs/((1+r)^tt)) - Cp/(1+r)^n
  Yi
}
Cs <- rep(1.34, times=20)
Cp <- 120
p <- 100
uniroot(Y_ma, c(0,1), p=p, Cs=Cs, Cp=Cp) 

```
We can get the yield to the maturity is **0.02151764**.

## 1.2 Macauley duration and modified duration

```{r}
du <- function(y,cupon,period,p0){
  c2<-0
  tc2<-0
  for(n in 1:period){
    t=n
    if(n<period) c<-cupon else if(n==period) c<-cupon+p0
    c1=c/(1+y)^n
    tc1<-t*c1
    c2<-c2+c1
    tc2<-tc2+tc1
    if(n==period) {
      d<-tc2/(c2*2)           
      md<-d/(1+y)              
    }
  }
  list(d=d,md=md)
}
(re <- du(0.0134,1.34,20,100))
```

The Macauley duration is **8.838122**.The Modified duration is **8.721257**.


 

## 1.3 Convexity
```{r}
td <- function(y, cupon, period, p0){
  c2<-0
  tc2<-0
  for(n in 1:period){
    t=n
    if(n<period) c<-cupon else if(n==period) c<-cupon+p0
    c1=c/(1+y)^n
    tc1<-t*(t+1)*c1
    c2<-c2+c1
    tc2<-tc2+tc1
    if(n==period){
      covex<-tc2/(c2*((1+y)^2))
      yearlycovex<-covex/4
    }
  }
  list(covex=covex,yearlycovex=yearlycovex)
}
(re <- td(0.0134,1.34,20,100))
# The market value of the debt change
delta.r <- -0.005
(effects <- 0.5*re$yearlycovex*delta.r^2)
```
The convexity is **346.2317**.The market value of the debt change is **0.001081974**.

# 2 The second question

## 2.1 Import data and convert it into a sequence object
I just took the first 100 days of the index.
```{r}
library(readxl)
library(dplyr)
library(knitr)
library(xts)  
library(forecast)
library(tseries)
library(sarima)

dat <- read_excel('D:/R/R-exercise/markdown-template/IDX_Idxtrd1.xlsx')%>%
  as.data.frame()

data <- dat[-c(1,2),-c(3:8)]

data$Idxtrd01 <- as.Date(data$Idxtrd01)

data$Idxtrd08 <- as.numeric(data$Idxtrd08)

data1 <- subset(data,Indexcd=='000001')
data1 <- data1[1:100,]

data2 <- subset(data,Indexcd=='000300')
data2 <- data2[1:100,]

data3 <- subset(data,Indexcd=='000802')
data3 <- data3[1:100,]

data.1 <- xts(data1[,3],as.Date(data1$Idxtrd01, format='%Y/%m/%d'))%>%
  as.ts()
data.2 <- xts(data2[,3],as.Date(data2$Idxtrd01, format='%Y/%m/%d'))%>%
  as.ts()
data.3 <- xts(data3[,3],as.Date(data1$Idxtrd01, format='%Y/%m/%d'))%>%
  as.ts()
```

## 2.2 Model identification

```{r}
par(mfrow=c(3,1)) 
plot(data.1)
plot(data.2)
plot(data.3)

adf.test(data.1)
adf.test(data.2)
adf.test(data.3)
ndiffs(data.1)

acf(data.1)
acf(data.2)
acf(data.3)
pacf(data.1)
pacf(data.2)
pacf(data.3)
```


From the above data, the trends of the three indices are almost the same.All three indexes passed the **ADF-test**, showing that the sequence is stationary.According to their ACF index and PACF index, white noise is featured.I am difficult to  judge q value and p value.Since the p+q value of ARMA model is as small as possible, I set p=q=0,and then gradually started from the low-order model to find the model with the lowest AIC value.Because the returns of the above three indexes tend to be the same, I will only use one of the indexes to build the model.

```{r}
am1 <- arima(data.1,order = c(0,0,0))
am1
am2 <- arima(data.1,order = c(0,0,1))
am2
am3 <- arima(data.1,order = c(1,0,0))
am3
am4 <- arima(data.1,order = c(1,0,1))
am4
am5 <- arima(data.1,order = c(2,0,1))
am5
am6 <- arima(data.1,order = c(1,0,2))
am6

```
We can get the message:

AIC(0,0,0)|AIC(0,0,1)|AIC(1,0,0)|AIC(1,0,1)|AIC(2,0,1)|AIC(1,0,2)|
|:-:|:-:|:-:|:-:|:-:|:-:|
365.27|367.25|367.25|369.19|368.19|368.4|
The smallest AIC value is 365.27.


Use **auto-arima()**  to help me select the model.

```{r}
am7 <- auto.arima(data.1)
am7
```
so,I choose ARIMA(0,0,0) as the best model.

## 2.3 Model diagnosis

White noise test and normal distribution test are performed for residuals.

```{r,fig.height= 4.5}
par(mfrow=c(1,1))
qqnorm(am1$residuals)
qqline(am1$residuals)
shapiro.test(am1$residuals)
```

```{r}
checkresiduals(am1)
```
The results show that the residual is white noise, but the residual is not normally distributed.

## 2.4 Model forecast
```{r,fig.height= 5}
autoplot(forecast(am1,5))
```
This time series is white noise. Because ARIMA(0,0,0) is a constant, the predicted value will also be a constant.
